Exploratory Data Analysis

### 2025-04-18 code/eda-code/eda.qmd
Work

This Quarto file loads the cleaned data and does some exploring.
# Setup

```{r}
#load needed packages. make sure they are installed
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
library(zoo)
library(tidyr)
library(tigris)
library(sf)
```

@@

# processed data
processeddata <- readRDS(here("data", "processed-data", "processeddata.rds"))

Creation of a heatmap showing total COVID 19 cases on the last observed date, 10/15/2022. I also created a heat map showing mobility changes, with hotter indicating more mobility to retail and recreational spots. Interestingly, the two maps seem to correlate even though I used total COVID-19 cases rather than new ones.
```{r}
# Load the state-level shapefile 
us_map_states <- tigris::states(cb = TRUE) %>%
  st_as_sf()  # Convert the shapefile to a Simple Feature object

# Clean the state names in the shapefile
us_map_clean <- us_map_states %>%
  mutate(NAME = tolower(NAME))  # Convert state abbreviation to lowercase

# Filter your dataset to get data for the last recorded day and group by state
last_day <- max(processeddata$date)
data_last_day <- processeddata %>%
  filter(date == last_day) %>%
  group_by(state) %>%
  summarise(total_cases = sum(total_cases, na.rm = TRUE),  # Sum of total cases
            retail_and_recreation_percent_change_from_baseline = 
              mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE)) %>%  # Average mobiltiy changes
  mutate(state = tolower(state))  # Convert state names to lowercase

# Join the shapefile with the COVID-19 data
us_map_data <- us_map_clean %>%
  left_join(data_last_day, by = c("NAME" = "state"))

# plot total COVID-19 cases to USA map on 10/15/2022
cases_map <- ggplot(us_map_data) +
  geom_sf(aes(fill = total_cases), color = "white") +
  scale_fill_viridis_c(option = "plasma", trans = "log", labels = NULL) + 
  labs(title = paste("Heatmap of Total COVID-19 Cases in the U.S. on", last_day),
       fill = "Hotter = More Cases") +
  theme_minimal() +
  coord_sf(xlim = c(-125, -65), ylim = c(25, 50))

print(cases_map)

figure_file = here("results","figures","casemap.png")
ggsave(filename = figure_file, plot=cases_map)

# Create the heatmap for retail and recreation mobility changes
mobility_map <- ggplot(us_map_data) +
  geom_sf(aes(fill = retail_and_recreation_percent_change_from_baseline), color = "white") +  
  scale_fill_viridis_c(option = "inferno", trans = "identity") +
  labs(title = paste("Heatmap of Retail and Recreation Mobility Changes in the U.S. on", last_day),
       fill = "Mobility Change") +
  theme_minimal() +
  theme(
        legend.spacing.y = unit(1, "cm"),  
        legend.box.spacing = unit(1, "cm"),  
        legend.key.size = unit(1, "cm")) +  
  coord_sf(xlim = c(-125, -65), ylim = c(25, 50))

# save graph
figure_file = here("results","figures","mobilitymap.png")
ggsave(filename = figure_file, plot=mobility_map)
```

This graph aims to plot total cases overtime for every location on a given day. This required the creation of a dataframe to store the sum of all total cases.
```{r}
# Create new data frame to have cases per day instead of state/county
total_cases_df <- processeddata %>%
  group_by(date) %>%
  summarise(Cases = sum(total_cases, na.rm = TRUE)) %>%
  ungroup()

# Plot the total case counts over time
case_count_graph <- ggplot(total_cases_df, aes(x = date, y = Cases)) +
  geom_line(color = "blue") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +  
  labs(title = "Total COVID-19 Cases Over Time in the U.S.",
       x = "Date",
       y = "Total Cases") +
  theme_minimal()

print(case_count_graph)

# save graph
figure_file = here("results","figures","casecountplot.png")
ggsave(filename = figure_file, plot=case_count_graph)
```
This graph aims to plot new cases overtime for every location on a given day. This required the creation of a dataframe to store the sum of all new cases.
```{r}
# data frame to sum all cases on a given day for every location
new_cases_df <- processeddata %>%
  group_by(date) %>%
  summarise(Cases = sum(new_cases, na.rm = TRUE)) %>%
  ungroup()

# Plot the new cases over time
incidence_graph <- ggplot(new_cases_df, aes(x = date, y = Cases)) +
  geom_line(color = "red") + 
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
  labs(title = "Daily New COVID-19 Cases Over Time in the U.S.",
       x = "Date",
       y = "New Cases per Day") +
  theme_minimal()

# save graph
figure_file = here("results","figures","incidenceplot.png")
ggsave(filename = figure_file, plot=incidence_graph)

```
I aim to plot out the changes in mobility for each category as time progresses. Originally, it plotted out everyday, however, the line was far too jagged so I applied a weekly rolling average instead. I also had to create a dataframe to average values for a given day for every location.
```{r}

# Create dataframe of average mobility for everywhere in the US per date
us_average_mobility <- processeddata %>%
  group_by(date) %>%
  summarise(
    total_mobility_changes = mean(average_mobility, na.rm = TRUE),
    transit_mean = mean(transit_stations_percent_change_from_baseline, na.rm = TRUE),
    parks_mean = mean(parks_percent_change_from_baseline, na.rm = TRUE),
    retail_mean = mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE),
    grocery_mean = mean(grocery_and_pharmacy_percent_change_from_baseline, na.rm = TRUE),
    workplaces_mean = mean(workplaces_percent_change_from_baseline, na.rm = TRUE),
    residential_mean = mean(residential_percent_change_from_baseline, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  arrange(date) %>%  # Ensure dates are sorted
  mutate(
    transit_mean_7day = rollmean(transit_mean, k = 7, fill = NA, align = "right"),
    parks_mean_7day = rollmean(parks_mean, k = 7, fill = NA, align = "right"),
    retail_mean_7day = rollmean(retail_mean, k = 7, fill = NA, align = "right"),
    grocery_mean_7day = rollmean(grocery_mean, k = 7, fill = NA, align = "right"),
    workplaces_mean_7day = rollmean(workplaces_mean, k = 7, fill = NA, align = "right"),
    residential_mean_7day = rollmean(residential_mean, k = 7, fill = NA, align = "right")
  )

# Plot the 7-day averages for each mobility type
mobility_graph <- ggplot(us_average_mobility, aes(x = date)) +
  geom_line(aes(y = grocery_mean_7day, color = "Grocery & Pharmacy")) +
  geom_line(aes(y = transit_mean_7day, color = "Transit Stations")) +
  geom_line(aes(y = workplaces_mean_7day, color = "Workplaces")) +
  geom_line(aes(y = residential_mean_7day, color = "Residential")) +
  geom_line(aes(y = parks_mean_7day, color = "Recreational")) +
  labs(
    title = "7-Day Moving Average of Percent Change in Mobility in the U.S. Over Time",
    x = "Date",
    y = "7-Day Average of Percent Change in Mobility",
    color = "Mobility Type"
  ) +
  theme_minimal()

@@

>>>>

This Quarto file loads the cleaned data and does some exploring. \# Setup

```{r}
#load needed packages. make sure they are installed
library(here) 
library(dplyr)
library(ggplot2)
library(zoo)
library(patchwork)
library(e1071)
library(scales)
library(RColorBrewer)
library(corrplot)
library(reshape2)

@@

# processed data
data <- readRDS(here("data", "processed-data", "processeddata.rds"))

case_data <- readRDS(here("data", "processed-data", "processedcasedata.rds"))

data<- data %>%
   mutate(incidence_rate = if_else(incidence_rate < 0, 0, incidence_rate))
```

DataVisualization

Since new case incidence is the variable of interest, it was visually inspected to understand the distribution.

From the graphs, kurtosis test, and skewness test, it is apparent that daily new case incidence is not normally distribued and heavily skewed towards 0 and other lower values. The second graph, new cases over 500, was graphed to show how certain days do reach high case incidence.
```{r}

density_cases <- data %>%
  ggplot( aes(x=new_cases)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
  labs(title = "Distribution of Daily New Case Incidence", 
       x = "Daily New Cases",
       y = "Density") +
  theme_bw()

print(density_cases)

figure_file = here("results","figures","density_cases.png")
ggsave(filename = figure_file, plot=density_cases)

density_cases_500 <- data %>%
  filter(new_cases > 500) %>%
  ggplot( aes(x=new_cases)) +
    geom_density(fill="#69b3a2", color="#e9ecef", alpha=0.8) +
  labs(title = "Distribution of Daily New Case Incidence (Over 500)", 
       x = "Daily New Cases",
       y = "Density") +
  theme_bw()

print(density_cases_500)

figure_file = here("results","figures","density_cases_500.png")
ggsave(filename = figure_file, plot=density_cases_500)

print(kurtosis(data$new_cases))
print(skewness(data$new_cases))
```
A histogram of population density was created. I hypothesize that population density plays a large role in case incidence. Population density is also skewed, however, there are some notable counties that do have large populationd densities.
```{r}
data_summary <- data %>%
  group_by(county) %>%
  summarize(pop_density = mean(pop_density, na.rm = TRUE))

histo_pop_density <- ggplot(data_summary, aes(x = pop_density)) + 
  geom_histogram(bins = 30, fill = "#69b3a2", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Population Density (2-Year Average per County)", 
       x = "Population Density (Calculated as Population Count / km squared)", y = "Frequency") + theme_bw()

print(histo_pop_density)

figure_file = here("results","figures","histo_pop_density.png")
ggsave(filename = figure_file, plot=histo_pop_density)
```

A 7-day rolling average was applied to the mobility data as the original iteration was too jagged. This code entails the creation of a data frame for 7 day averages.

# Create dataframe of average mobility per county per date
ga_average_mobility <- data %>%
  group_by(date) %>%
  summarise(
    retail_mean = mean(retail_and_recreation_percent_change_from_baseline, na.rm = TRUE),
    grocery_mean = mean(grocery_and_pharmacy_percent_change_from_baseline, na.rm = TRUE),
    workplaces_mean = mean(workplaces_percent_change_from_baseline, na.rm = TRUE),
    residential_mean = mean(residential_percent_change_from_baseline, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  arrange(date) %>%  # Ensure dates are sorted
  mutate(
    retail_mean_7day = rollmean(retail_mean, k = 7, fill = NA, align = "right"),
    grocery_mean_7day = rollmean(grocery_mean, k = 7, fill = NA, align = "right"),
    workplaces_mean_7day = rollmean(workplaces_mean, k = 7, fill = NA, align = "right"),
    residential_mean_7day = rollmean(residential_mean, k = 7, fill = NA, align = "right")
  )

# Plot the 7-day averages for each mobility type
mobility_graph <- ggplot(ga_average_mobility, aes(x = date)) +
  geom_line(aes(y = grocery_mean_7day, color = "Grocery & Pharmacy")) +
  geom_line(aes(y = workplaces_mean_7day, color = "Workplaces")) +
  geom_line(aes(y = residential_mean_7day, color = "Residential")) +
  geom_line(aes(y = retail_mean_7day, color = "Retail")) +
  labs(
    title = "7-Day Moving Average of Percent Change in Mobility in Georgia Over Time",
    x = "Date",
    y = "7-Day Average of Percent Change in Mobility",
    color = "Mobility Type"
  ) +
  theme_bw()

@@

For better legiiblity, all 4 mobiltiy types were graphed separately.

```{r}
# Create one plot per mobility type
p1 <- ggplot(ga_average_mobility, aes(x = date, y = grocery_mean_7day)) +
  geom_line(color = "steelblue") +
  labs(title = "Grocery & Pharmacy", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

p2 <- ggplot(ga_average_mobility, aes(x = date, y = workplaces_mean_7day)) +
  geom_line(color = "darkred") +
  labs(title = "Workplaces", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

p3 <- ggplot(ga_average_mobility, aes(x = date, y = residential_mean_7day)) +
  geom_line(color = "forestgreen") +
  labs(title = "Residential", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

p4 <- ggplot(ga_average_mobility, aes(x = date, y = retail_mean_7day)) +
  geom_line(color = "purple") +
  labs(title = "Retail", x = "Date", y = "7-Day Avg % Change") +
  theme_bw() +
  coord_cartesian(ylim = c(-50, 20))

# Arrange in a 2x2 grid
mobility_grid <- (p1 | p2) / (p3 | p4)

# Show it
print(mobility_grid)

# Save it
figure_file = here("results", "figures", "mobility_grid.png")
ggsave(filename = figure_file, plot = mobility_grid)

After visually inspecting the new cases over time graph, heteroscedastity and non-stationairty are of concern. Large outbreaks can skew these metrics and even when tested, could be misleading. Additionally, assumptions of homoscedastity are required for some tests of stationairty and some assumptions of normality are required for some tests of homoscdastity. Because of these concerns, I chose to focus on models like XGBoost and Random Forest which do not require these assumptions.

```{r}
library(scales)

first_date <- min(data$date)
last_date <- max(data$date)

cases_graph <- case_data %>%
  group_by(date) %>%
  summarise(total_cases = sum(total_cases, na.rm = TRUE)) %>%
  arrange(date) %>%
  ggplot(aes(x = date, y = total_cases)) +
    geom_area(fill = "#69b3a2", alpha = 0.5) +
    geom_line(color = "#69b3a2") +
    ylab("Total Cases (in GA)") +
    scale_y_continuous(labels = label_number()) +
   scale_x_date(
      breaks = seq(from = first_date, to = last_date, by = "6 months"),  # Breaks every 3 months
      labels = date_format("%b %Y")  # Format labels as month year (e.g., Jan 2020)
    ) +
    theme_bw() +
  labs(title = "Total Cases (in GA) Over Time",
       x = "Date (2020 - 2022)",
       y = "Total Cases")

print(cases_graph)

figure_file = here("results","figures","cases_graph.png")
ggsave(filename = figure_file, plot=cases_graph)

incidence_graph <- data %>%
  group_by(date) %>%
  summarise(new_cases = sum(new_cases, na.rm = TRUE)) %>%
  arrange(date) %>%
  ggplot(aes(x = date, y = new_cases)) +
    geom_area(fill = "#69b3a2", alpha = 0.5) +
    geom_line(color = "#69b3a2") +
    ylab("Daily New Cases") +
     scale_x_date(
      date_breaks = "3 months",  # Automatically set the breaks to every 3 months
      date_labels = "%b %Y"      # Format labels as month-year (e.g., Jan 2020)
    ) +
    theme_bw() +
  labs(title = "Daily New Cases (in GA) Over Time",
       x = "Date (2020 - 2022)",
       y = "Total Cases")

figure_file = here("results","figures","incidence_graph.png")
ggsave(filename = figure_file, plot=incidence_graph)
```

This code chunk plots a scatter plot of a county's population count against its final total cases. 
```{r}

year_end_cases <- data %>%
  mutate(year = year(date)) %>%
  group_by(county, year) %>%
  slice_max(order_by = date, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(county, year, total_cases)

pop_density_info <- data %>%
  mutate(year = year(date)) %>%
  select(county, year, pop_density) %>%
  distinct()

plot_df <- left_join(year_end_cases, pop_density_info, by = c("county", "year"))

year_end_cases_graph <- ggplot(plot_df, aes(x = pop_density, y = total_cases)) +
  geom_point(color = "#69b3a2", alpha = 0.6) +
  facet_wrap(~year) +
  labs(
    x = "Population Density",
    y = "Total COVID Cases (End of Year)",
    title = "Year-End COVID Cases vs. Population Density by County"
  ) +
  theme_bw()

print(year_end_cases_graph)

figure_file = here("results","figures","year_end_cases_graph.png")
ggsave(filename = figure_file, plot=year_end_cases_graph)
```
Correlation Analysis

This code chunk entails correlation testing through Spearman's and the creation of a matrix. First, lagged variables for case count was created for the data.
```{r}

data <- data %>%
  group_by(county) %>%
  arrange(date) %>%
  mutate(
    lag_1 = lag(new_cases, 1),
    lag_7 = lag(new_cases, 7),
    lag_14 = lag(new_cases, 14)
  ) %>%
  ungroup()

data <- data %>% drop_na()

# Create a data frame for matrix creation
cor_df <- data %>%
  select("New Cases" = new_cases, "1 Day Lag" = lag_1, "7 Day Lag" = lag_7, "14 Day Lag" = lag_14, "Population Density" = pop_density, "Retail Mobility" = retail_and_recreation_percent_change_from_baseline, "Workplace Mobility" = workplaces_percent_change_from_baseline, "Residential Mobility" = residential_percent_change_from_baseline, "Grocery Mobility" = grocery_and_pharmacy_percent_change_from_baseline,)

# Calculate correlations using Spearman's rank correlation because data is non parametric
cor_matrix <- cor(cor_df, method = "spearman")

# Melt for visualization
cor_matrix_melted <- melt(cor_matrix)

# Create plot for matrix
matrix <- ggplot(cor_matrix_melted, aes(x = Var1, y = Var2)) +
  geom_point(aes(size = abs(value), color = value)) +
  geom_text(aes(label = round(value, 2)), color = "black", size = 3) +
  scale_color_gradient2(low = "#BFA2DB", high = "#92C7CF", mid = "white",
                        midpoint = 0, limit = c(-1, 1), name = "Correlation") +
  scale_size(range = c(2, 10), guide = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap (Spearman's)", x = "Variables", y = "Variables")

print(matrix)

figure_file = here("results","figures","matrix.png")
ggsave(filename = figure_file, plot=matrix)
```
